---
title: "Big Data Analysis Final Project - Coral Reef Fisheries in Indonesia"
author: "Paul Carvalho"
date: "May 1, 2017"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

Overfishing, climate change, coastal development, and pollution are deteriorating marine
ecosystems and threatening the welfare of coastal communities (1, 2). This problem is
developing rapidly on coral reefs where growing human populations and increasing demand for
seafood are fueling overfishing and compromising the ecosystem services provided by coral
reefs (3). Marine Protected Areas (MPAs) that ban fishing activities have been shown to provide
conservation benefits by protecting biodiversity and ecological functions that make coral reefs
resilient (4). Although MPAs are touted by conservationists and adopted as the primary tool for
ecosystem-based management (EBM), MPAs are not a panacea for marine resource management issues
(5). For example, in coral reef systems there can be little community support for conservation
management that drastically changes cultural traditions, often making MPAs socially
unacceptable (6). Consequently, gear-based restrictions may be an alternative tool for EBM that
can be tailored to the local, sociocultural context of communities with coral reef fisheries, and thus
generating more community support and promoting conservation of coral reef habitats and
associated ecosystem services (7).

Many gears are used in coral reef fisheries around the world, such as hook-and-line,
spearguns, traps, beach seines, and gill nets (8). Due to the diversity in fishing methods and
gears, gear-types differentially select species, trophic levels, and size classes of fish (9).
Understanding the effects of different gears on reef fish assemblages is important as many
species perform salient roles in the ecosystem (10). For example, herbivorous fish graze turf or
macroalgae that compete with corals for resources (i.e., space), effectively preventing shifts from
coral-dominated to algal-dominated habitats (11). If certain gears target and remove large
quantities of herbivores, habitat phase shifts occur, and coral reefs lose their structural
complexity and ability to support productive fisheries, an important ecosystem service (12).
Also, certain gears (i.e., gill nets, beach seines) can be damaging to coral reef habitats and reduce
structural complexity via entanglement (13). In light of this, it is critical to gain a better
understanding of the ecological impacts of fishing gears in coral reef systems if effective EBM is
to be developed.For this project I analyze catch landings data from coral reef fisheries in Lombok, Indonesia. I aim to analyze the species composition and size selectivity of different fishing gears used in Lombok. 

## Methods

In 2014 and 2015, the Wildlife Conservation Society Indonesia Program (WCS) collected fishery landings data from Lombok, Indonesia located in the Nusa Tenggara Barat region (Fig. 1). WCS personnel randomly sampled fishsers as they returned from a fishing trip. Data collectors identified fish to the species level and weight the total catch. In addition, they estimated average weight (kg) and size (cm) of individual fish.

I used R software to clean data and remove unnecessary variables from the data. Also, I checked for misspelling of all variables in the data and created a new file with cleaned data. After data cleaning, I analyzed data using R Software.

```{r, include=FALSE, echo=FALSE}
# Indonesia map libraries ---------------------------------------------------------------------------------------
library(maps)
library(mapdata)
```

```{r, echo=FALSE}
cc <- map('world', names = TRUE, plot = FALSE)
take <- unlist(sapply(c("indonesia"), 
                      grep, tolower(cc), value = TRUE))
map('world', regions=take, fill=TRUE, col='grey93')
map(regions="indonesia", fill=TRUE, col='grey45', add=TRUE)
points(116.3249, -8.6510, col = "red", cex = 3, lwd = 3)
box(which="plot", lty="solid")
```
**Figure 1.** Map of Indonesia with Lomnok indicated by red circle.

## Data Analysis

Handlines caught the most fish in terms of individuals caught and total weight (Fig. 2). Although spearguns caught much less fish in weight, spearguns and handlines caught similar quantities of individual fish (Fig. 3). Due to insufficient data collection for fish caught by nets, I excluded net data from the following data analyses. Scarce data from net fisheries might be due to failure to sample net fishers, or from the infrequent use of nets in Lombok, or both.

```{r, include=FALSE, echo=FALSE}
# Directories ---------------------------------------------------------------------------------------
# setwd("C:/Users/pgcar/Desktop/BDA/Project")
setwd("C:/Users/humphriesLAB/Desktop/Paul Carvalho/Courses/Big Data/BDA_Project")

# Libraries -----------------------------------------------------------------------------------------
library(plyr)
library(dplyr)
library(stringr)
library(ggplot2)
library(plotrix)

# Functions -----------------------------------------------------------------------------------------
calc_percent_total = function(catch_data, by_weight){
  # Calculate the percent of catch for each row in the data frame.
  # 
  # Input:
  #   catch_data - a dataframe for catch data.
  #   by_weight - TRUE for weight and FALSE for frequency 
  #
  # Output:
  #   Returns a vector with percentages of total weight or frequency (all values sum to 100).
  if(by_weight==TRUE){
    percent_catch = (catch_data$total_catch / (sum(catch_data$total_catch, na.rm=TRUE))) * 100
    return(percent_catch)
  }
  else{
    percent_frequency = (catch_data$total_frequency / (sum(catch_data$total_frequency, na.rm=TRUE))) * 100
    return(percent_frequency)
  }
  
}

sort_catch = function(catch_data, by_weight, sort_decreasing){
  # Sort the levels in factor "Genus_Species" by percent catch or frequency.
  #
  # Input:
  #   catch_data - a dataframe for catch data.
  #   by_weight - TRUE to sort by weight and FALSE to sort by freuency.
  #   sort_decreasing - If TRUE sort percent catch in decreasing order. If FALSE sort in increasing order.
  #
  # Output:
  #   Returns the sorted factor "Genus_Species".
  if(by_weight == TRUE){
    return(factor(catch_data$Genus_Species, 
                  levels = catch_data$Genus_Species[order(catch_data$percent_catch,decreasing=sort_decreasing)]))
  }
  else{
    return(factor(catch_data$Genus_Species, 
                  levels = catch_data$Genus_Species[order(catch_data$total_frequency,decreasing=sort_decreasing)]))
  }
  
}


# Load data -----------------------------------------------------------------------------------------
raw_lombok_data = read.csv("Lombok_cleaned.csv", header = TRUE)
# Create a variable with genus and species name
raw_lombok_data$Genus_Species = paste(raw_lombok_data$Genus, raw_lombok_data$Species, sep = " ")
raw_lombok_data$Frequency = raw_lombok_data$Total_catch_kg/raw_lombok_data$Weight_per_fish_kg

```

```{r, echo=FALSE}
# Total catch by gear type --------------------------------------------------------------------------
data_geartype = ddply(raw_lombok_data, c("Fishing_Gear_Group"), 
                      function(df)c(total_catch=sum(df$Total_catch_kg,na.rm=TRUE),
                                    total_frequency=sum(df$Frequency,na.rm=TRUE)))
# Remove NA
data_geartype = data_geartype[-4,]
# Plot catch for each gear type by weight
ggplot(data_geartype, aes(x=Fishing_Gear_Group, y=total_catch)) + 
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle=0))+
  labs(x="Fishing Gear", y="Total fish weight (kg)")
```
**Figure 2**. Total catch (kilograms) by all fishing gears in Lombok, Indonesia.


```{r, echo=FALSE}
# Plot catch for each gear type by frequency
ggplot(data_geartype, aes(x=Fishing_Gear_Group, y=total_frequency)) + 
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle=0))+
  labs(x="Fishing Gear", y="Number of fish")
```
**Figure 3**. Total number of fish caught by all fishing gears in Lombok, Indonesia.  

```{r, include=FALSE, echo=FALSE}
# Catch composition: entire catch -------------------------------------------------------------------
# Sum of total catch for each species and create new dataframe
lombok_data_no_net = subset(raw_lombok_data, Fishing_Gear_Group != "Net") # Remove net data
lombok_data_no_net$Fishing_Gear_Group = factor(lombok_data_no_net$Fishing_Gear_Group) # Remove net from factor levels
total_catch_df = ddply(lombok_data_no_net, c("Genus_Species"), 
                       function(df)c(total_catch=sum(df$Total_catch_kg,na.rm=TRUE),
                                     total_frequency=sum(df$Frequency,na.rm=TRUE)))
# Remove the first row because it does not contain data
total_catch_df = total_catch_df[-1,]

# Check for NAs
num_NA = sum(is.na(total_catch_df))

# Calculate percent of total weight and frequency for each species
total_catch_df$percent_catch = calc_percent_total(total_catch_df, TRUE)
total_catch_df$percent_frequency = calc_percent_total(total_catch_df, FALSE)

# Sort by weight and frequency
sorted_total_weight = total_catch_df[order(total_catch_df$percent_catch,decreasing = TRUE),]
sorted_total_frequency = total_catch_df[order(total_catch_df$total_frequency,decreasing = TRUE),]

# Sort the order of levels in factor genus_species for sorted dataframes
sorted_total_weight$Genus_Species = sort_catch(sorted_total_weight,TRUE,TRUE)
sorted_total_frequency$Genus_Species = sort_catch(sorted_total_frequency,FALSE,TRUE)
```

In total, WCS collected catch data for 194 species of reef fish, but most species make up less than 1% of the total catch data. The timor snapper (*Lutjanus timoriensis*) accounts for over 20% of the total catch by weight, followed by a small rockfish (*Sebasticus tertius*) and the elongate surgeonfish (*Acanthurus mata*) (Fig.4). The same three species comprise the top species in terms of number of individuals caught. However, the eolongate surgeonfish makes up most of the catch (17%), and the timor snapper is third at ~5% of the total catch (Fig.5).

```{r, echo=FALSE}
# Plot catch composition by weight (top 20 species)
ggplot(sorted_total_weight[1:20,], aes(x=Genus_Species, y=percent_catch)) + 
  geom_bar(stat='identity') +
  ylab(expression(atop("Percent of total catch", paste("(by weight)")))) +
  labs(x="Species") +
  theme(axis.text.x = element_text(angle=60, hjust=1, vjust=1)) +
  theme(axis.line.x = element_line(color='black', size=0.5),
        axis.line.y = element_line(color='black', size=0.5))
```
**Figure 4**. Species catch composition by weight for handlines and spearguns in Lombok, Indonesia. This figure only includes the top 20 species by weight.

```{r, echo=FALSE}
# Plot catch composition by frequency (top 20 species)
ggplot(sorted_total_frequency[1:20,], aes(x=Genus_Species, y=percent_frequency)) + 
  geom_bar(stat='identity') +
  ylab(expression(atop("Percent of total catch", paste("(by indivisuals)")))) +
  labs(x="Species") +
  theme(axis.text.x = element_text(angle=60, hjust=1, vjust=1)) +
  theme(axis.line.x = element_line(color='black', size=0.5),
        axis.line.y = element_line(color='black', size=0.5))
```
**Figure 5**. Species catch composition by frequency (i.e., number of individuals caught) for handlines and spearguns in Lombok, Indonesia. This figure only includes the top 20 species by individuals caught.

```{r, include=FALSE, echo=FALSE}
# Subset data by fishing gear -----------------------------------------------------------------------
gear_catch_df = ddply(lombok_data_no_net, c("Genus_Species","Fishing_Gear_Group"), 
                       function(df)c(total_catch=sum(df$Total_catch_kg),
                                     total_frequency=sum(df$Frequency)))
# Spears
spear_df = subset(gear_catch_df, Fishing_Gear_Group == 'Speargun')
# Handline
handline_df = subset(gear_catch_df, Fishing_Gear_Group == 'Handline')
```

```{r, include=FALSE, echo=FALSE}
# Spear ---------------------------------------------------------------------------------------------
# Calculate percent total catch
spear_df$percent_catch = calc_percent_total(spear_df,TRUE)
spear_df$percent_frequency = calc_percent_total(spear_df,FALSE)

# Sort by weight and frequency
spear_sorted_weight = spear_df[order(spear_df$percent_catch,decreasing = TRUE),]
spear_sorted_frequency = spear_df[order(spear_df$total_frequency,decreasing = TRUE),]

# Sort the order of levels in factor genus_species for sorted dataframes
spear_sorted_weight$Genus_Species = sort_catch(spear_sorted_weight,TRUE,TRUE)
spear_sorted_frequency$Genus_Species = sort_catch(spear_sorted_frequency,FALSE,TRUE)
```

```{r, include=FALSE, echo=FALSE}
# Handline ---------------------------------------------------------------------------------------------
# Calculate percent total catch
handline_df$percent_catch = calc_percent_total(handline_df,TRUE)
handline_df$percent_frequency = calc_percent_total(handline_df,FALSE)

# Sort by weight and frequency
handline_sorted_weight = handline_df[order(handline_df$percent_catch,decreasing = TRUE),]
handline_sorted_frequency = handline_df[order(handline_df$total_frequency,decreasing = TRUE),]

# Sort the order of levels in factor genus_species for sorted dataframes
handline_sorted_weight$Genus_Species = sort_catch(handline_sorted_weight,TRUE,TRUE)
handline_sorted_frequency$Genus_Species = sort_catch(handline_sorted_frequency,FALSE,TRUE)
```

When catch separated by fishing gear type, I found that fishers using spearguns mainly target the elongate surgeonfish, *Acanthurus mata* (Fig. 6 and 7). 

```{r, echo=FALSE}
# Spear plots ----------------------------------------------------------------------------------------------
# Plot top 20 of the catch by weight
ggplot(spear_sorted_weight[1:20,], aes(x=Genus_Species, y=percent_catch)) + 
  geom_bar(stat='identity') +
  labs(x="Species", y="Percent of total catch") +
  theme(axis.text.x = element_text(angle=60, hjust=1, vjust=1)) +
  theme(axis.line.x = element_line(color='black', size=0.5),
        axis.line.y = element_line(color='black', size=0.5))
```
**Figure 6**. Species catch composition by weight for speargun catch in Lombok, Indonesia.

```{r, echo=FALSE}
# Plot top 20 of the catch by frequency
ggplot(spear_sorted_frequency[1:20,], aes(x=Genus_Species, y=percent_frequency)) + 
  geom_bar(stat='identity') +
  labs(x="Species", y="Percent of total catch") +
  theme(axis.text.x = element_text(angle=60, hjust=1, vjust=1)) +
  theme(axis.line.x = element_line(color='black', size=0.5),
        axis.line.y = element_line(color='black', size=0.5))
```
**Figure 7**. Species catch composition by frequency for speargun catch in Lombok, Indonesia.

```{r, echo=FALSE}
# Handline plots ----------------------------------------------------------------------------------------------
# Plot top 20 of the catch by weight
ggplot(handline_sorted_weight[1:20,], aes(x=Genus_Species, y=percent_catch)) + 
  geom_bar(stat='identity') +
  labs(x="Species", y="Percent of total catch") +
  theme(axis.text.x = element_text(angle=60, hjust=1, vjust=1)) +
  theme(axis.line.x = element_line(color='black', size=0.5),
        axis.line.y = element_line(color='black', size=0.5))
```
**Figure 8**. Species catch composition by weight for handline catch in Lombok, Indonesia.

```{r, echo=FALSE}
# Plot top 20 of the catch by frequency
ggplot(handline_sorted_frequency[1:20,], aes(x=Genus_Species, y=percent_frequency)) + 
  geom_bar(stat='identity') +
  labs(x="Species", y="Percent of total catch") +
  theme(axis.text.x = element_text(angle=60, hjust=1, vjust=1)) +
  theme(axis.line.x = element_line(color='black', size=0.5),
        axis.line.y = element_line(color='black', size=0.5))
```
**Figure 9**. Species catch composition by frequency for handline catch in Lombok, Indonesia
